---
title: "MBTA Calculations"
author: "Marshall Thomas"
date: "September 30, 2017"
output: github_document
---

## Summary

- For this analysis, I used data available from the MBTA API.
- I compared July-September of 2016 to the same time in 2017.
- I focused on rush hour traffic (6:00 AM - 10:00 AM; 4:00 PM - 8:00 PM).
- I used the longest possible trip queries for each line. Practically, this 
meant querying from the first station to the second-to-last station, so these 
transit times are one station short of end-to-end. For example, Ashmont
northbound is from Ashmont to Davis; while the Orange line southbound is from
Oak Grove to Green Street.

```{r include=FALSE}
# Load libraries
library(lubridate)
library(ggplot2)
library(dplyr)
library(gridExtra)
options(stringsAsFactors = FALSE)

#make ggplot2 better
theme_basic = theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", size = 1),
        axis.ticks.length = unit(0.2, "lines"),
        axis.text.x = element_text(margin=margin(10,5,5,5,"pt")),
        axis.text.y = element_text(margin=margin(5,10,5,5,"pt")))
theme_set(theme_basic)
```

```{r echo=FALSE}
# Orange line southbound (oak grove to green street)
orange_SB = read.csv("OrangeLineSouthBound")

# Orange line northbound (forest hills to malden center)
orange_NB = read.csv("OrangeLineNorthBound")

# Blue line westbound 
blue_WB = read.csv("BlueLineWestBound")

# Blue line eastbound
blue_EB = read.csv("BlueLineEastBound")

# Red line southbound to Ashmont (Alewife to Shawmut)
red_SB_Ash = read.csv("AshmontSouthBound")

# Red line southbound to Braintree (Alewife to Quincy Adams)
red_SB_Brain = read.csv("BraintreeSouthBound")

# Red line northbound from Ashmont (Ashmont to Davis)
red_NB_Ash = read.csv("AshmontNorthBound")

# Red line northbound from Braintree (Braintree to Davis)
red_NB_Brain = read.csv("BraintreeNorthBound")
```

```{r echo=FALSE}
# Function that grabs only rush hour data
rush_hourify = function(dataset){
  dataset$weekdays = dataset$day != "Saturday" & dataset$day != "Sunday"
  dataset$rushtime = dataset$time > "06:00:00" & dataset$time < "10:00:00"
  dataset$morning_peak = dataset$rushtime & dataset$weekdays
  dataset$PMrush = dataset$time > "16:00:00" & dataset$time < "20:00:00" 
  dataset$afternoon_peak = dataset$PMrush & dataset$weekdays
  dataset = subset(dataset, dataset$morning_peak | dataset$afternoon_peak)
  return(dataset)}

# Function that subsets data for June - September
month_subset = function(dataset){
  dataset$Months2017 = dataset$just_date > "2017-06-30" & 
  dataset$just_date < "2017-10-01"
  dataset$Months2016 = dataset$just_date > "2016-06-30" & 
  dataset$just_date < "2016-10-01"

  datanew = subset(dataset, (dataset$Months2017 | dataset$Months2016))
  datanew$range = ifelse(datanew$Months2017, "July-Sept 2017", "July-Sept 2016")
  return(datanew)
}

# Runs the major processing steps of the data
processor = function(dataset){
  step1 = rush_hourify(dataset)
  step1$delay = step1$travel_times > (step1$benchmark_times + 300)
  step2 = month_subset(step1)
  return(step2)
}

# Sets a time cutoff for the delays, calculates a number of delays relative to MBTA benchmark
delaycalc = function(dataset, delaytime){
  dataset$delayed = dataset$travel_times > (dataset$benchmark_times + delaytime)
  delays = dataset %>%
  group_by(just_date) %>% 
  summarise (ndelays = sum(delayed))
  delays$daterange = ifelse(delays$just_date > "2017-01-01","July-Sept 2017","July-Sept 2016")
  return(delays)
}

# Subsets on morning peak, and finds the number of trains that run each morning
AMtrainscalc = function(dataset){
  dataset_morning = subset(dataset, dataset$morning_peak)
  Ntrains = dataset_morning %>% group_by(just_date) %>% 
    summarise (counts = length(travel_times))
  Ntrains$daterange = ifelse(Ntrains$just_date > "2017-01-01",
                             "July-Sept 2017","July-Sept 2016")
  return(Ntrains)
}

```


```{r echo=FALSE}
# Plotting functions

# Violin plot of travel time
violinplot = function(dataset, title){
  plot = ggplot(data = dataset, aes(x = range, y = travel_times)) +
  geom_violin() +
  ylab("Travel time (seconds)") +
  xlab("Date range") +
  ggtitle(title)
  return(plot)
}

# Boxplot for number of delays
boxplot = function(dataset, title){
  plot = ggplot(data = dataset, aes(x = daterange, y = ndelays)) +
  geom_boxplot() +
  ylab("Number of delays (>5 min) per day") +
  xlab("Date range") +
  ggtitle(title)
  return(plot)
}

# Boxplot for number of trains
ntrainsplot = function(dataset, title){
  plot = ggplot(data = dataset, aes(x = daterange, y = counts)) +
  geom_boxplot() +
  ylab("Number of trains per morning") +
  xlab("Date range") +
  ggtitle(title)
  return(plot)
}
  
```

```{r echo=FALSE}
# Set up the delays table
Alldelays = read.csv(text="Summer2016,Summer2017")

# The delays function
delay_add = function(line_data, delays_table,desc){
  NDelay = line_data %>% group_by(daterange) %>% 
  summarise (mean = mean(ndelays))
NDelay_Line = t(NDelay$mean)
colnames(NDelay_Line) = names(delays_table)
rownames(NDelay_Line) = c(desc)
delays_table = rbind(delays_table, NDelay_Line)
return(delays_table)
}

# Set up the train counts table
Allcounts = read.csv(text="Summer2016,Summer2017")

# The counts function
count_add = function(line_data, counts_table,desc){
  NTrains = line_data %>% group_by(daterange) %>%
    summarise (mean = mean(counts))
NTrains_Line = t(NTrains$mean)
colnames(NTrains_Line) = names(counts_table)
rownames(NTrains_Line) = c(desc)
counts_table = rbind(counts_table, NTrains_Line)
return(counts_table)
}
```

```{r echo=FALSE}
### Data for Orange Line Southbound (Oak Grove to Green St)

orange_SB_JulySept = processor(orange_SB)
orange_SB_delays = delaycalc(orange_SB_JulySept, 300)
Alldelays = delay_add(orange_SB_delays,Alldelays,"Orange line southbound")
orange_SB_ntrainsAM = AMtrainscalc(orange_SB_JulySept)
Allcounts = count_add(orange_SB_ntrainsAM,Allcounts,"Orange line southbound")

# Boxplot here
delay_OrangeSB = boxplot(orange_SB_delays, "Orange line southbound")
```

```{r echo=FALSE}
### Data for Orange Line Northbound (Forest Hills to Malden Center)

orange_NB_JulySept = processor(orange_NB)
orange_NB_delays = delaycalc(orange_NB_JulySept, 300)
Alldelays = delay_add(orange_NB_delays,Alldelays,"Orange line northbound")
orange_NB_ntrainsAM = AMtrainscalc(orange_NB_JulySept)
Allcounts = count_add(orange_NB_ntrainsAM,Allcounts,"Orange line northbound")

# Boxplot here
delay_OrangeNB = boxplot(orange_NB_delays, "Orange line northbound")
```

```{r echo=FALSE}
### Data for Blue Line Westbound (Wonderland to Government Center)

blue_WB_JulySept = processor(blue_WB)
blue_WB_delays = delaycalc(blue_WB_JulySept, 300)
Alldelays = delay_add(blue_WB_delays,Alldelays,"Blue line westbound")
blue_WB_ntrainsAM = AMtrainscalc(blue_WB_JulySept)
Allcounts = count_add(blue_WB_ntrainsAM,Allcounts,"Blue line westbound")

# Boxplot here
delay_BlueWB = boxplot(blue_WB_delays, "Blue line westbound")
```

```{r echo=FALSE}
### Data for Blue Line Eastbound (Bowdoin to Revere Beach)

blue_EB_JulySept = processor(blue_EB)
blue_EB_delays = delaycalc(blue_EB_JulySept, 300)
Alldelays = delay_add(blue_EB_delays,Alldelays,"Blue line eastbound")
blue_EB_ntrainsAM = AMtrainscalc(blue_EB_JulySept)
Allcounts = count_add(blue_EB_ntrainsAM,Allcounts,"Blue line eastbound")

# Boxplot here
delay_BlueEB = boxplot(blue_EB_delays, "Blue line eastbound")
```

```{r echo=FALSE}
### Data for Ashmont Red Line Southbound (Alewife to Shawmut)

red_SB_Ash_JulySept = processor(red_SB_Ash)
red_SB_Ash_delays = delaycalc(red_SB_Ash_JulySept, 300)
Alldelays = delay_add(red_SB_Ash_delays,Alldelays,"Ashmont southbound")
red_SB_Ash_ntrainsAM = AMtrainscalc(red_SB_Ash_JulySept)
Allcounts = count_add(red_SB_Ash_ntrainsAM,Allcounts,"Ashmont southbound")

# Boxplot here
delay_RedSB_Ash = boxplot(red_SB_Ash_delays, "Ashmont southbound")
```

```{r echo=FALSE}
### Data for Red Line Southbound (Alewife to Quincy Adams)

red_SB_Brain_JulySept = processor(red_SB_Brain)
red_SB_Brain_delays = delaycalc(red_SB_Brain_JulySept, 300)
Alldelays = delay_add(red_SB_Brain_delays,Alldelays,"Braintree southbound")
red_SB_Brain_ntrainsAM = AMtrainscalc(red_SB_Brain_JulySept)
Allcounts = count_add(red_SB_Brain_ntrainsAM,Allcounts,"Braintree southbound")

# Boxplot here
delay_RedSB_Brain = boxplot(red_SB_Brain_delays, "Braintree southbound")
```

```{r echo=FALSE}
### Data for Red Line Northbound (Ashmont to Davis)

red_NB_Ash_JulySept = processor(red_NB_Ash)
red_NB_Ash_delays = delaycalc(red_NB_Ash_JulySept, 300)
Alldelays = delay_add(red_NB_Ash_delays,Alldelays,"Ashmont northbound")
red_NB_Ash_ntrainsAM = AMtrainscalc(red_NB_Ash_JulySept)
Allcounts = count_add(red_NB_Ash_ntrainsAM,Allcounts,"Ashmont northbound")

# Boxplot here
delay_RedNB_Ash = boxplot(red_NB_Ash_delays, "Ashmont northbound")
```

```{r echo=FALSE}
### Data for Red Line Northbound (Braintree to Davis)

red_NB_Brain_JulySept = processor(red_NB_Brain)
red_NB_Brain_delays = delaycalc(red_NB_Brain_JulySept, 300)
Alldelays = delay_add(red_NB_Brain_delays,Alldelays,"Braintree northbound")
red_NB_Brain_ntrainsAM = AMtrainscalc(red_NB_Brain_JulySept)
Allcounts = count_add(red_NB_Brain_ntrainsAM,Allcounts,"Braintree northbound")

# Boxplot here
delay_RedNB_Brain = boxplot(red_NB_Brain_delays, "Braintree northbound")
```

### Delays by train line

- Summer2016 covers July - September 2016
- Summer2017 covers July - September 2017
- Delays are for rush hours (6:00 - 10:00 AM and 4:00 - 8:00 PM).
- Delays are relative to the MBTA "benchmark" time given for that trip.
- **Conclusion: substantially increased delays since last summer on the orange line.**
```{r echo=FALSE}
Alldelays
```

### Number of trains running by train line

- Summer2016 covers July - September 2016
- Summer2017 covers July - September 2017
- Number of trains during morning rush hour only (6:00 - 10:00 AM).
- **Conclusion: fewer morning trains running on the orange line since last summer (though not by much).**
```{r echo=FALSE}
Allcounts
```

### Figure showing delays on each train line
```{r figure1, fig.height=12, fig.width=7, echo=FALSE}
grid.arrange(delay_BlueEB, delay_BlueWB, delay_OrangeSB, delay_OrangeNB, 
             delay_RedSB_Ash, delay_RedNB_Ash, 
             delay_RedSB_Brain, delay_RedNB_Brain,
             top = "Delays of more than 5 minutes", ncol=2)
```

### Orange line delays deep dive

- The average number of delays per day this summer is **roughly equivalent to
the number of delays on the worst day last summer**.
- There are fewer trains running during the morning rush hour on the orange line.
- There has been a significant increase in trip times on the orange line.
```{r figure2, fig.height=10, fig.width=7, echo=FALSE}
# Plot orange SB
violin_OrangeSB = violinplot(orange_SB_JulySept, "Orange line southbound")

# Plot orange NB
violin_OrangeNB = violinplot(orange_NB_JulySept, "Orange line northbound")

trains_SB_orange = ntrainsplot(orange_SB_ntrainsAM, "Orange line southbound")
trains_NB_orange = ntrainsplot(orange_NB_ntrainsAM, "Orange line northbound")

grid.arrange(delay_OrangeSB, delay_OrangeNB, 
             trains_SB_orange, trains_NB_orange,
             violin_OrangeSB, violin_OrangeNB,
             top = "Summary of orange line data (peak times only)", ncol=2)
```

