---
title: "Orange Line Dwell Times"
author: "Marshall Thomas"
date: "9/30/2017"
output: github_document
---

```{r include=FALSE}
# Load libraries
library(jsonlite)
library(lubridate)
library(ggplot2)
library(dplyr)
options(stringsAsFactors = FALSE)

#make ggplot2 better
theme_basic = theme_bw() +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_blank(),
        panel.border = element_rect(colour = "black", size = 1),
        axis.ticks.length = unit(0.2, "lines"),
        axis.text.x = element_text(margin=margin(10,5,5,5,"pt")),
        axis.text.y = element_text(margin=margin(5,10,5,5,"pt")))
theme_set(theme_basic)

```

```{r echo=FALSE}
# Function for queries from data API
Dwell_query = function(platform,start_datetime,finish_datetime){
  start_time = as.integer(as.POSIXct((start_datetime), tz="America/New_York"))
  finish_time = as.integer(as.POSIXct((finish_datetime), tz="America/New_York"))
  url_start = "http://realtime.mbta.com/developer/api/v2.1/dwells?api_key="
  key = "wX9NwuHnZU2ToO7GmGR9uw"
  url_int = "&to_datetime="
  arrival_times = c()
  dwell_times = c()

  while(start_time<finish_time){
    end = start_time + 86400
    request_url = paste(url_start, key, "&format=json&stop=", platform,
    "&from_datetime=", start_time, "&to_datetime=", end, sep = "")
    dat = fromJSON(request_url)
    arrival = as.numeric(dat$dwell_times$arr_dt)
    times = as.numeric(dat$dwell_times$dwell_time_sec)
    dwell_times = c(dwell_times, times)
    arrival_times = c(arrival_times, arrival)
    start_time = end
  }

  date = as.POSIXct(arrival_times, tz ="America/New_York", origin ="1970-01-01")
  just_date = as.Date(date, tz ="America/New_York")
  time = strftime(date, tz="America/New_York", format="%H:%M:%S")
  day = weekdays(just_date)
  compiled_times = cbind.data.frame(date,day,just_date,time,dwell_times)
  return(compiled_times)
}
```

```{r echo = FALSE}
# Function that grabs only morning rush hour data
rush_hourify = function(dataset){
  dataset$weekdays = dataset$day != "Saturday" & dataset$day != "Sunday"
  dataset$rushtime = dataset$time > "06:00:00" & dataset$time < "10:00:00"
  dataset$morning_peak = dataset$rushtime & dataset$weekdays
  dataset$PMrush = dataset$time > "16:00:00" & dataset$time < "20:00:00" 
  dataset$afternoon_peak = dataset$PMrush & dataset$weekdays
  dataset = subset(dataset, dataset$morning_peak | dataset$afternoon_peak)
  return(dataset)}
```


```{r echo = FALSE}
#Queries parameters for stations


get_dwells = function(Nstop){
  start17 = "2017-07-01 00:00:00"
  end17 = "2017-10-01 00:00:00"
  start16 = "2016-07-01 00:00:00"
  end16 = "2016-10-01 00:00:00"
  train_inbound_2017 = Dwell_query(Nstop, start17, end17)
  train_inbound_2016 = Dwell_query(Nstop, start16, end16)
  train_inbound = rbind(train_inbound_2016,train_inbound_2017)
  train_inbound$stop_number = Nstop
  train_inbound$range = ifelse(train_inbound$just_date > "2017-01-10", 
                                  "July-Sept 2017", "July-Sept 2016")
  train_inbound_RH = rush_hourify(train_inbound)  
  return(train_inbound_RH)
}
```

```{r echo = FALSE}
# Query all inbound stations (up to downtown crossing) on orange line southbound
run_dwell_queries = function(){
  alldwells = get_dwells("70036")
  alldwells = rbind(alldwells, get_dwells("70034"))
  alldwells = rbind(alldwells, get_dwells("70032"))
  alldwells = rbind(alldwells, get_dwells("70278"))
  alldwells = rbind(alldwells, get_dwells("70030"))
  alldwells = rbind(alldwells, get_dwells("70028"))
  alldwells = rbind(alldwells, get_dwells("70026"))
  alldwells = rbind(alldwells, get_dwells("70024"))
  alldwells = rbind(alldwells, get_dwells("70022"))
  return(alldwells)
}
```

```{r echo = FALSE}
# Load data from a .csv, otherwise, run the query and save the data
if(file.exists("inbound_dwell_times")){
  dwells_inbound = read.csv("inbound_dwell_times")
} else{
  dwells_inbound = run_dwell_queries
}
```


## Summary

Here, I analyzed the dwell times on the southbound branch of the orange line,
 from Malden Center to State Street. I hypothesized that the delays in 2017 
 could be due to overcrowding. If this were the case, we would have longer platform 
 delays, because when there is a crowd on the platform it generally takes longer
 for the  doors to close. I planned to compare crowding on the southbound
 inbound  trains in the morning and afternoon, because  delays in the inbound 
 trains could be due to anything (mechanical failures, crowding, etc.), but 
 crowding would be much worse in the morning, when most commuters are heading 
 towards the city. Thus, if crowding were the cause of the slower service in 
 2017, we would expect to see longer dwell times in the morning inbound commute 
 in 2017 relative to 2016. However, there is no major difference in dwell times.
 
```{r echo = FALSE}
# Calculate mean dwell times in the morning and afternoon, in 2016 and 2017
mean_dwells = dwells_inbound %>%
  group_by(range, afternoon_peak, stop_number) %>% 
  summarise (mediandwell = median(dwell_times), meandwell = mean(dwell_times))

dwells_morning = subset(dwells_inbound, dwells_inbound$morning_peak & 
                          dwells_inbound$stop_number !="70036")
dwells_afternoon = subset(dwells_inbound, dwells_inbound$afternoon_peak & 
                          dwells_inbound$stop_number !="70036")

morning_dwells = dwells_morning %>%
  group_by(range) %>% 
  summarise (mediandwell = median(dwell_times), meandwell = mean(dwell_times))

afternoon_dwells = dwells_afternoon %>%
  group_by(range) %>% 
  summarise (mediandwell = median(dwell_times), meandwell = mean(dwell_times))

```

### Morning dwell times - 2016 vs. 2017

- This includes all stations from Malden Center to State Street
- Oak Grove was excluded because the trains wait much longer there (however, 
including Oak Grove does not affect the conclusion of the analysis)
- This is focused on the morning peak (6:00 - 10:00 AM)
- This compares July-Sept 2016 with July-Sept 2017

```{r echo = FALSE}
# Plot the average dwell times in the morning
ggplot(dwells_morning, aes(x = range, y = dwell_times)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2", breaks = c(16, 64, 256, 1024, 4096), 
                     limits = c(12,4096)) +
  ylab("Dwell time (seconds)") +
  xlab("Date range") +
  ggtitle("Orange line southbound - morning rush")
```

### Afternoon dwell times - 2016 vs. 2017

- This includes all stations from Malden Center to State Street
- Oak Grove was excluded because the trains wait much longer there (however, 
including Oak Grove does not affect the conclusion of the analysis)
- This is focused on the afternoon peak (4:00 - 8:00 PM)
- This compares July-Sept 2016 with July-Sept 2017

```{r echo = FALSE}
# Plot the average dwell times in the afternoon
ggplot(dwells_afternoon, aes(x = range, y = dwell_times)) +
  geom_boxplot() +
  scale_y_continuous(trans = "log2", breaks = c(16, 64, 256, 1024, 4096), 
                     limits = c(12,4096)) +
  ylab("Dwell time (seconds)") +
  xlab("Date range") +
  ggtitle("Orange line southbound - afternoon rush")
```

